### 2024-07-16 journal/algorithm.tex
cleaning up some .tex files (removing comments, \rev{})

%\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ performs the following
three steps.
$P$ first observes
the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
\rev{Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does} this three-step sequence a
\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on~$\var{v}$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $\var{v}$'s children to $\var{v}$,
we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
then $P$'s operation has been propagated to $\var{v}$ by some other process.
\rev{Thus, after performing a double \op{Refresh},} $P$ can continue 
onwards towards the~root.

@@

If a block contained an explicit \rev{listing} of its sequences of \enqueues\ and \dequeues,
it would take $\Omega(p)$ \rev{steps} to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{orderingtree} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

%The information stored in a \block\ depends on whether it is in an internal node or a leaf.
A block in a leaf represents a single \enqueue\ or \dequeue.  The block's \fld{element} field stores the value
enqueued if the operation is an \enqueue, or \nl\ if the operation is a \dequeue.

Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child, \rev{respectively}.  
Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\hspace*{3mm}
\end{eqnarray}
The \eleft\ and \eright\ fields allow us to navigate to a block's direct subblocks.
Blocks also store some prefix sums:
%the block in 
$\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total numbers of \enqueues\ and \dequeues, \rev{respectively}, in $\var{v}.\fld{blocks}[1..\var{b}]$.
We use these to search for a particular operation.
% pinpoint the  location of an operation among the subblocks of a given block.
For example, consider finding the $r$th \enqueue\ $E_r$ in the linearization.
A binary search for $r$ on the \fld{sum\sub{enq}} fields of the root's blocks 
finds the block  containing $E_r$.
If we know a block $B$ in a \node\ $\var{v}$ contains $E_r$,
we can use the \fld{sum\sub{enq}} field again to determine which child of $\var{v}$ contains $E_r$
and then do a binary search
of the \fld{sum\sub{enq}} fields of the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find the leaf block that  stores 
$E_r$ explicitly.

\rev{Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.}
We shall show \rev{in Section \ref{sec-analysis}} that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
%All information needed for this search process can be derived from the 
%\eleft, \eright\ and \fld{sum\sub{enq}} fields.

A block is called the \emph{superblock} of all of its direct subblocks.
To facilitate navigating upwards in the ordering tree for task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
%We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation  up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ \rev{steps}.

@@

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ \rev{steps}.
A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute 
\begin{eqnarray*}
B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
the number of \enqueues\ and \dequeues\ in~$B$
\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
The \enqueues\ of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
Then the \dequeues\ of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have}
\begin{eqnarray}
B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\nonumber\\ 
			&& \hspace*{10mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).\label{size-eqn}
\end{eqnarray}

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $\var{v}$, we do not yet know where its
superblock will eventually be installed in \var{v}'s parent.
So, we leave $B.\fld{super}$ blank.  
Soon after $B$ is installed,
%a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by 
some process will
set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
We shall show that this happens soon enough \rev{to ensure} that $B.\fld{super}$ can differ from the true index of \rev{$B$'s superblock}
by at most 1.
%After $B$ is installed 
%in $\var{v}.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
%when they attempt to advance $\var{v}.\fld{head}$ from $h$ to $h+1$,
%using a value they read from the \fld{head} field of $\var{v}$'s parent.
%As mentioned above, this might not be the exact index of $B$'s superblock, but we
%shall prove that it is close.

@@

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figures \ref{code1}--\ref{code3}.
%We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
%$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
%the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
%We use \var{num\sub{deq}} similarly.
%\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf.}\fld{blocks} and propagate to root}}\label{enqAppend}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf.}\fld{blocks} and propagate to root}}\label{deqAppend}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root.}\fld{blocks}}}\label{invokeIndexDequeue}
    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \dequeue\ should return}}\label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  \rev{if first \op{Refresh} fails, try again}}
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \Comment{\rev{help advance $v$'s children's \var{head} indices, if necessary}}\label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}}\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlock}
    \If{$\var{B} = \nl$} \Return{\tr}\Comment{\rev{if no new operations to propagate to $v$, do nothing}}\label{addOP} 
	\Else
	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}}\Comment{\rev{try to add $B$ to \var{v}.\fld{blocks}}}\label{cas}
    	\State \Call{Advance}{\var{v, h}} \Comment{\rev{advance $v.head$ from $h$ to $h+1$}}\label{advance}
    	\State \Return{ \var{result}}\Comment{\rev{return true iff \op{CAS} succeeded}}
	\EndIf
\EndFunction{Refresh}
\end{algorithmic}
\caption{Queue implementation's main routines.\label{code1}}
\end{figure*}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} 
    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th \dequeue\ of $D(\var{root}.\fld{blocks}[\var{b}'])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} \dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else \Comment{first, find the superblock of $v.\fld{blocks}[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\Comment{\rev{actual index of superblock is either \var{sup} or $\var{sup}+1$}}\label{idsup1}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
	    \EndIf\label{idsup3}
	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock}
	    \State \linecomment{\rev{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# \dequeues\ in superblock from left sibling}}
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}} \Comment{\rev{recurse up the tree}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th \dequeue\ in $D(\var{root}.\fld{blocks}[\var{b}])$
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
    %  \State \linecomment{contains at least \var{i} dequeues}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \Comment{queue is empty when \dequeue\ occurs}\label{returnNull}
    \Else \ \Comment{response is the \var{e}th \enqueue\ in the root}
        \State \var{e} \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
		\State use \rev{doubling} search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of \enqueue\ within its block}\label{FRi}
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that \enqueue}}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}
\end{algorithmic}
\caption{Queue implementation's helper routines. \rev{Recall that $D(B)$ is the sequence of \dequeues\ represented by block $B$.}\label{code2}}
\end{figure*}

@@

%Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
%As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
%We shall show {$B$.\fld{super}}  differs
%from the index of $B'$ by at most~1.

>>>>

To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ performs the following
three steps.
$P$ first observes
the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does this three-step sequence a
\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on~$\var{v}$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $\var{v}$'s children to $\var{v}$,
we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
then $P$'s operation has been propagated to $\var{v}$ by some other process.
Thus, after performing a double \op{Refresh}, $P$ can continue 
onwards towards the~root.

@@

If a block contained an explicit listing of its sequences of \enqueues\ and \dequeues,
it would take $\Omega(p)$ steps to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{orderingtree} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

A block in a leaf represents a single \enqueue\ or \dequeue.  The block's \fld{element} field stores the value
enqueued if the operation is an \enqueue, or \nl\ if the operation is a \dequeue.

Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child, respectively.  
Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\hspace*{3mm}
\end{eqnarray}
The \eleft\ and \eright\ fields allow us to navigate to a block's direct subblocks.
Blocks also store some prefix sums:
$\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total numbers of \enqueues\ and \dequeues, respectively, in $\var{v}.\fld{blocks}[1..\var{b}]$.
We use these to search for a particular operation.
For example, consider finding the $r$th \enqueue\ $E_r$ in the linearization.
A binary search for $r$ on the \fld{sum\sub{enq}} fields of the root's blocks 
finds the block  containing $E_r$.
If we know a block $B$ in a \node\ $\var{v}$ contains $E_r$,
we can use the \fld{sum\sub{enq}} field again to determine which child of $\var{v}$ contains $E_r$
and then do a binary search
of the \fld{sum\sub{enq}} fields of the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find the leaf block that  stores 
$E_r$ explicitly.

Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.
We shall show in Section \ref{sec-analysis} that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.

A block is called the \emph{superblock} of all of its direct subblocks.
To facilitate navigating upwards in the ordering tree for task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation  up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ steps.

@@

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ steps.
A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute 
\begin{eqnarray*}
B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
the number of \enqueues\ and \dequeues\ in~$B$
as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
The \enqueues\ of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
Then the \dequeues\ of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have
\begin{eqnarray}
B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\nonumber\\ 
			&& \hspace*{10mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).\label{size-eqn}
\end{eqnarray}

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $\var{v}$, we do not yet know where its
superblock will eventually be installed in \var{v}'s parent.
So, we leave $B.\fld{super}$ blank.  
Soon after $B$ is installed,
some process will
set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
We shall show that this happens soon enough to ensure that $B.\fld{super}$ can differ from the true index of 
$B$'s superblock by at most 1.
%After $B$ is installed 
%in $\var{v}.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
%when they attempt to advance $\var{v}.\fld{head}$ from $h$ to $h+1$,
%using a value they read from the \fld{head} field of $\var{v}$'s parent.
%As mentioned above, this might not be the exact index of $B$'s superblock, but we
%shall prove that it is close.

@@

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figures \ref{code1}--\ref{code3}.

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
    \State \Call{Append}{\var{B}}\Comment{append \op{Enqueue} to \var{leaf.}\fld{blocks} and propagate to root}\label{enqAppend}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
    \State \Call{Append}{\var{B}}\Comment{append \dequeue\ to \var{leaf.}\fld{blocks} and propagate to root}\label{deqAppend}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{find where \op{Deqeueue} is in \var{root.}\fld{blocks}}\label{invokeIndexDequeue}
    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{find response \dequeue\ should return}\label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  if first \op{Refresh} fails, try again}
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \Comment{help advance $v$'s children's \var{head} indices, if necessary}\label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}}\Comment{create new block to append to \var{v}.\fld{blocks}}\label{invokeCreateBlock}
    \If{$\var{B} = \nl$} \Return{\tr}\Comment{if no new operations to propagate to $v$, do nothing}\label{addOP} 
	\Else
	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}}\Comment{try to add $B$ to \var{v}.\fld{blocks}}\label{cas}
    	\State \Call{Advance}{\var{v, h}} \Comment{advance $v.head$ from $h$ to $h+1$}\label{advance}
    	\State \Return{ \var{result}}\Comment{return true iff \op{CAS} succeeded}
	\EndIf
\EndFunction{Refresh}
\end{algorithmic}
\caption{Queue implementation's main routines.\label{code1}}
\end{figure*}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} 
    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th \dequeue\ of $D(\var{root}.\fld{blocks}[\var{b}'])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} \dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else \Comment{first, find the superblock of $v.\fld{blocks}[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\Comment{actual index of superblock is either \var{sup} or $\var{sup}+1$}\label{idsup1}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
	    \EndIf\label{idsup3}
	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock}
	    \State \linecomment{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
        \If{$\fld{dir} = \fld{right}$} \Comment{add \# \dequeues\ in superblock from left sibling}
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}} \Comment{recurse up the tree}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th \dequeue\ in $D(\var{root}.\fld{blocks}[\var{b}])$
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \Comment{queue is empty when \dequeue\ occurs}\label{returnNull}
    \Else \ \Comment{response is the \var{e}th \enqueue\ in the root}
        \State \var{e} \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
		\State use doubling search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of \enqueue\ within its block}\label{FRi}
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{find argument of that \enqueue}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}
\end{algorithmic}
\caption{Queue implementation's helper routines. Recall that $D(B)$ is the sequence of \dequeues\ represented by block $B$.\label{code2}}
\end{figure*}

@@

To compute the response of the $i$th \dequeue\ in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If not, line \ref{computeE} computes the rank \var{e} of the
\op{Enqueue} whose argument is the \dequeue's response. 
The routine next searches for
the index $b_e$ of the block in \var{root}.\fld{blocks} that contains 
the \var{e}th \enqueue.
Since the \enqueue\ is linearized before the \dequeue, \linebreak $b_e\leq b$.  To find the left end of the range to search for $b_e$, \op{FindResponse} does a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, the doubling search performs a binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} within that range to find $b_e$.
Finally, the call to \op{GetEnqueue} at line \ref{findAnswer} traces down through the tree to find the required \enqueue\ in a leaf.

\opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
$i$th \enqueue\ in the $b$th block $B$ of \typ{Node} $\var{v}$. 
It recursively finds the location of the \enqueue\ in each node along the path from $v$ to a leaf, which stores the argument explicitly.
The test at line \ref{leftOrRight} determines which child of \var{v} contains
the \enqueue\ by checking whether $i$ is less than or equal to the number of  
\enqueues\ in $B$ that came from \var{v}'s left child.
Line \ref{getChild} determines the range of blocks within that child that are subblocks of $B$
using information stored
in $B$ and the block that precedes $B$ in $v$,
and then
finds the exact subblock containing the \enqueue\ using a binary search on the \fld{sum\sub{enq}}
field of the child's \var{blocks} array.
Line \ref{getChildIndex} computes the index of the \enqueue\ within the block
in the child, and line \ref{getRecurse}
proceeds recursively down the tree.
